Burning sage: Reversing the curse of dimensionality in the visualisation of high-dimensional data

### 2020-09-22 paper/paper.Rmd
fig 3 reversed and tweaked  explanation

```{r, out.width="60%", fig.cap="Illustration (and notation) for describing the element used in the burning sage transformation. The 3D sphere (left) shows the different volumes to be compared. The full sphere has volume $V(R, p)$. Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red. The intersection of the plane with the sphere is illustrated in grey, and the plane representing the projection with both radii is shown at right.", fig.align = "center"}

@@

To understand how the $p$ dimensional volume is projected onto a $2$ dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius. This dependence was described in @Laa:2020wkm. We start from a $p$ dimensional hypersphere with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r; p, R)$, where $r$ can be any radius within $[0, R]$. The relative projected volume is then given as the ratio of these two quantities,
\begin{equation}
v_{2} (r; p, R) = \frac{V_{2D}(r; p, R)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation}
This ratio is of particular interest because it gives the 2-dimensional radial cumulative distribution function (CDF) of points when assuming a uniform distribution within the $p$-dimensional hypersphere.

We can compare $v_{2} (r; p, R)$ to the relative volume within a radius $r$ in the original $p$-dimensional hypersphere,
\begin{equation}
v_{p} (r; p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
\end{equation}
This comparison is shown in Fig. \ref{fig:cdf}, for $p=3, 10, 100$. The left graph shows $v_{p} (r; p, R)$ and the right graph $v_{2} (r; p, R)$. We clearly see the opposing behavior of the two functions as $p$ increases. While $v_{p} (r; p, R)$ concentrates more and more towards large values of $r/R$, $v_{2} (r; p, R)$ has the inverse behavior.

```{r cdf, fig.cap="Comparing the relative volume of a $p$ dimensional hypersphere captured within a radius $r$, in the $p$-dimensional space (left) and in a 2-dimensional projection (right), for $p=3, 10, 100$. While most of the volume is squeezed towards larger values of $r$ as $p$ increases, the projected volume gets concentrated near the center instead.", fig.height=2.2, fig.width=6, out.width = "90%", fig.align = "center"}

@@

grid.arrange(p1, p2, ncol=2)

>>>>

```{r, out.width="60%", fig.cap="Illustration (and notation) for describing the elements used in the burning sage transformation. The 3D sphere (left) shows the different volumes to be compared. The full sphere has volume $V(R, p)$. Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red. The intersection of the plane with the sphere is illustrated in grey, and the plane representing the projection with both radii is shown at right.", fig.align = "center"}

@@

To understand how the $p$-dimensional volume is projected onto a $2$ dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius. This dependence was described in @Laa:2020wkm. We start from a $p$-dimensional hypersphere, with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r; p, R)$, where $r$ can be any radius within $[0, R]$. The relative projected volume is then given as the ratio of these two quantities,
\begin{equation}
v_{2} (r; p, R) = \frac{V_{2D}(r; p, R)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation}
This ratio is of particular interest because it gives the 2-dimensional radial cumulative distribution function (CDF) of points when assuming a uniform distribution within the $p$-dimensional hypersphere.

We can compare $v_{2} (r; p, R)$ to the relative volume within a radius $r$ in the original $p$-dimensional hypersphere,
\begin{equation}
v_{p} (r; p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
\label{eq:vp}
\end{equation}

Figure \ref{fig:cdf} compares these two quantities (equations \ref{eq:cdf} and \ref{eq:vp}), for $p=3, 10, 100$. At left is $v_{2} (r; p, R)$ and on the right is $v_{p} (r; p, R)$. The function shapes change in opposite directions as $p$ increases: $v_{2} (r; p, R)$ peaks earlier and $v_{p} (r; p, R)$ gets flatter. This is the paradox of the curse of dimensionality in that projected volume at the center increases with $p$.

```{r cdf, fig.cap="Comparing relative volume of a $p$-dimensional hypersphere captured within a radius $r$, in the 2-dimensional space (left) and in a $p$-dimensional projection (right), for $p=3, 10, 100$. The difference is dramatic, which illustrates the paradox of the curse of dimensionality. The relative volume of a $p$D sphere shrinks, as $p$ increases, while the projected volume (near the center) grows.", fig.height=2.2, fig.width=6, out.width = "90%", fig.align = "center"}

@@

grid.arrange(p2, p1, ncol=2)

### 2020-09-22 paper/paper.tex
fig 3 reversed and tweaked  explanation

\caption{Illustration (and notation) for describing the element used in the burning sage transformation. The 3D sphere (left) shows the different volumes to be compared. The full sphere has volume $V(R, p)$. Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red. The intersection of the plane with the sphere is illustrated in grey, and the plane representing the projection with both radii is shown at right.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

To understand how the \(p\) dimensional volume is projected onto a \(2\)
dimensional plane, we study what fraction of the total volume is
projected onto the area of a disk depending on its radius. This
dependence was described in \citet{Laa:2020wkm}. We start from a \(p\)
dimensional hypersphere with radius \(R\) and volume \(V(R, p)\), and
its projected volume onto a centered \(2\)-dimensional disk of radius
\(r\), \(V_{2D}(r; p, R)\), where \(r\) can be any radius within
\([0, R]\). The relative projected volume is then given as the ratio of
these two quantities, \begin{equation}
v_{2} (r; p, R) = \frac{V_{2D}(r; p, R)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation} This ratio is of particular interest because it gives the
2-dimensional radial cumulative distribution function (CDF) of points
when assuming a uniform distribution within the \(p\)-dimensional
hypersphere.

We can compare \(v_{2} (r; p, R)\) to the relative volume within a
radius \(r\) in the original \(p\)-dimensional hypersphere,
\begin{equation}
v_{p} (r; p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
\end{equation} This comparison is shown in Fig. \ref{fig:cdf}, for
\(p=3, 10, 100\). The left graph shows \(v_{p} (r; p, R)\) and the right
graph \(v_{2} (r; p, R)\). We clearly see the opposing behavior of the
two functions as \(p\) increases. While \(v_{p} (r; p, R)\) concentrates
more and more towards large values of \(r/R\), \(v_{2} (r; p, R)\) has
the inverse behavior.

@@

\caption{Comparing the relative volume of a $p$ dimensional hypersphere captured within a radius $r$, in the $p$-dimensional space (left) and in a 2-dimensional projection (right), for $p=3, 10, 100$. While most of the volume is squeezed towards larger values of $r$ as $p$ increases, the projected volume gets concentrated near the center instead.}\label{fig:cdf}
\end{figure}

>>>>

\caption{Illustration (and notation) for describing the elements used in the burning sage transformation. The 3D sphere (left) shows the different volumes to be compared. The full sphere has volume $V(R, p)$. Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red. The intersection of the plane with the sphere is illustrated in grey, and the plane representing the projection with both radii is shown at right.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

To understand how the \(p\)-dimensional volume is projected onto a \(2\)
dimensional plane, we study what fraction of the total volume is
projected onto the area of a disk depending on its radius. This
dependence was described in \citet{Laa:2020wkm}. We start from a
\(p\)-dimensional hypersphere, with radius \(R\) and volume \(V(R, p)\),
and its projected volume onto a centered \(2\)-dimensional disk of
radius \(r\), \(V_{2D}(r; p, R)\), where \(r\) can be any radius within
\([0, R]\). The relative projected volume is then given as the ratio of
these two quantities, \begin{equation}
v_{2} (r; p, R) = \frac{V_{2D}(r; p, R)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation} This ratio is of particular interest because it gives the
2-dimensional radial cumulative distribution function (CDF) of points
when assuming a uniform distribution within the \(p\)-dimensional
hypersphere.

We can compare \(v_{2} (r; p, R)\) to the relative volume within a
radius \(r\) in the original \(p\)-dimensional hypersphere,
\begin{equation}
v_{p} (r; p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
\label{eq:vp}
\end{equation}

Figure \ref{fig:cdf} compares these two quantities (equations
\ref{eq:cdf} and \ref{eq:vp}), for \(p=3, 10, 100\). At left is
\(v_{2} (r; p, R)\) and on the right is \(v_{p} (r; p, R)\). The
function shapes change in opposite directions as \(p\) increases:
\(v_{2} (r; p, R)\) peaks earlier and \(v_{p} (r; p, R)\) gets flatter.
This is the paradox of the curse of dimensionality in that projected
volume at the center increases with \(p\).

@@

\caption{Comparing relative volume of a $p$-dimensional hypersphere captured within a radius $r$, in the 2-dimensional space (left) and in a $p$-dimensional projection (right), for $p=3, 10, 100$. The difference is dramatic, which illustrates the paradox of the curse of dimensionality. The relative volume of a $p$D sphere shrinks, as $p$ increases, while the projected volume (near the center) grows.}\label{fig:cdf}
\end{figure}
